---
title: "Importing Data Files using stringr and purrr"
author: "Alice Tivarovsky"
date: '2020-09-05'
toc: true
slug: import-function
categories:
  - Programming
  - R
tags:
  - Programming
  - R
---

## Motivation

It's a well-known [tenet](https://en.wikipedia.org/wiki/Rule_of_three_(computer_programming)) of computer programming that if you have to copy and paste code more than twice, you should write a function.  

In my [phthalates exposure projects](/blog/2020-05-30-phthalates), I imported 21 text files for tidying and analysis. This is clearly a task well-suited for a function, but instead, I tediously copied and pasted code 21 times. Why? Honestly, I just didn't have much experience writing functions and I was too eager to get data into R to actually sit down and figure out how to automate it. 

I knew, though, that one day, I'd sit down and figure out how to write that function. Today was that day. 

A few things to note: 
1. I'm working with two file types here, and neither one is in a common .csv format. I downloaded these files from [NHANES](https://wwwn.cdc.gov/nchs/nhanes/default.aspx), which plays nicely with SAS, but less nicely with R. Some of the raw data they provide is in .txt format and some of it is in the SAS-ready .xport format. I'm using the `read.xport` function in the `foreign` library to import both types, but the function can easily be extended to more common file types. 
2. The function below is specific to the situation, wherein the files are downloaded not only in .xport/.txt format, but they are named using the years first and survey type second. This led to error messages when I tried to save the imported dataframes using their original filenames (R doesn't like objects named with numbers first), so I did some character re-shuffling using a few `stringr` functions. If your files are named more appropriately, you won't need this feature.  

## Code

As mentioned above, we'll need the `foreign` library to read the xport files. We'll also need `stringr` for some of the naming manipulations we'll be doing within the function. 
```{r global_options, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")
```


```{r}
library(tidyverse)
library(foreign)
library(stringr)

```

First, we create a vector of filenames from the data folder and check its length.
```{r}
files = list.files("./data/phthalates")
files
length(files)

```

Next, we write a function that does the following: 
1. Reads in a vector of names and concatenates it with the folder path, resulting in a vector of file path names called `y`
2. Reads in the dataset represented by filepath `y`
3. Removes the file extensions from y, since we'll want these strings to name our imported dataframes
4. Extracts the years, which originally come first in our filenames vector - otherwise, R will throw errors when we try to access the dataframes
5. Rearranges the strings so that the years are at the end
6. Stores the imported datasets, assigning a name using the resulting string from steps 3-5. 
```{r function}
mult_import = function(x){
  y = str_c("./data/phthalates/", x) 
  z = as.data.frame(read.xport(y))
  name_1 = str_replace_all(x, ".txt|.XPT", "")
  years = str_sub(x, 1, 9)
  name_2 = str_sub(name_1, 11, 20)
  new_name = str_c(name_2, years, sep = "_")
  assign(new_name, z, envir = .GlobalEnv)
}

```

Next, we use `purrr` to apply `mult_import` to our datafiles. And voila! We have 21 neat dataframes to work with. I'm only going to print 3 of them here (just to prove that it worked). 
```{r purrr}
data_sets = map(files, mult_import)

head(DEMO_C_2003_2004)
head(DBQ_G_2011_2012)
head(PHTHTE_C_2003_2004)

```

## Conclusion

Writing these 10 lines of code required some time and trial and error. Certainly, copy/pasting code and modifying snippets would have been faster. But that's a lesson learned in the process of automation in code, which usually follows the so-called [flywheel effect](https://www.jimcollins.com/concepts/the-flywheel.html). It might take some effort to get the thing spinning, but once it picks up enough momentum, you get smooth movement with minimal additional input.  

