---
title: "Adding Subset Buttons to plotly Plots"
author: "Alice Tivarovsky"
date: "2022-09-18"
toc: true
slug: plotly-buttons
output: blogdown::html_page
tags:
  - R
  - Data Visualization
  - Epidemiology
categories:
  - R
  - Data Visualization
  - Epidemiology
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE, 
  fig.align = "center", 
  out.width = "100%"
  )

```


## Motivation

I recently found myself drafting some `plotly` figures for work, and I wanted to add a bit more interactivity without creating a `shiny` app. I plotted time-series data and I wanted the end user to be able to subset the plot to a year of their choosing, or to see the entire range of time in one figure.

After a bit of googling around, I found a few solutions, but [this one](https://plotly.com/r/custom-buttons/) suited my needs the best. These buttons are straight-forward for the user and pretty simple to code up. Below is a quick demo. 


## Data Preparation

```{r libraries}
library(tidyverse)
library(plotly)

```


We'll be using Covid-19 case data in the United States, downloaded from [Our World in Data](https://ourworldindata.org/explorers/coronavirus-data-explorer?facet=none&Metric=Confirmed+cases&Interval=7-day+rolling+average&Relative+to+Population=true&Color+by+test+positivity=false&country=~USA). 

```{r load data}
raw_data <- 
  read.csv("./data/owid-covid-data.csv", header = TRUE) 

head(raw_data)

```

Next, we'll limit to United States data points, and select only the necessary columns. We'll use `new_cases_smoothed` (smoothed over the week of data) to avoid the noise of daily fluctuation around the line. We'll also fix the variable class of `date`. 

```{r clean data}
us_df <- 
  raw_data %>% 
  select(location, date, new_cases_smoothed) %>% 
  filter(location == "United States") %>% 
  mutate(date = as.POSIXct(date))

us_df %>% 
  filter(!is.na(new_cases_smoothed)) %>% 
  head()

# huge file, remove to save memory
rm(raw_data)

```

## Analysis

We will start by plotting a regular `plotly` figure, as a "before" shot, if you will. 

```{r regular plot}
regular_plot <- 
  us_df %>% 
  plot_ly(x = ~date) %>% 
  add_lines(y = ~new_cases_smoothed) %>% 
  layout(title = "Covid-19 Incidence in the US", 
         yaxis = list(title = "New Cases (Smoothed)"))

regular_plot

```
<br>
Now, we'll apply the buttons. To do this, we will use an updatemenu `relayout` method, as detailed [here](https://plotly.com/r/custom-buttons/). There are several methods, but `relayout` is the one most appropriate, since it will allow the user to modify a layout attribute, in this case the range of the x-axis.  

```{r plot with buttons}
button_plot <- 
  us_df %>% 
  plot_ly(x = ~date) %>% 
  add_lines(y = ~new_cases_smoothed) %>% 
  layout(
    title = "Covid-19 Incidence in the US", 
    yaxis = list(title = "New Cases (Smoothed)"),
    updatemenus = list(
      list(type = "buttons",
           buttons = list(list(label = "All", 
                              method = "relayout", 
                              args = list(list(xaxis = list(range = c("2020-01-01","2022-12-31"))))),
                           list(label = "2020", 
                                method = "relayout", 
                                args = list(list(xaxis = list(range = c("2020-01-01","2020-12-31"))))),
                          list(label = "2021", 
                                method = "relayout", 
                                args = list(list(xaxis = list(range = c("2021-01-01","2021-12-31"))))),
                          list(label = "2022", 
                                method = "relayout", 
                                args = list(list(xaxis = list(range = c("2022-01-01","2022-12-31")))))
                          ))))

button_plot
  

```

<br>

## Conclusion

And that's it! We took a `plotly` figure and made it even more functional with a few easy-to-code controls. Just when I thought I knew the ins and outs of `plotly`, it surprised me with a sweet feature I've never seen.
